---
title: "stats2data_application"
output:
  rmarkdown::html_vignette:
    df_print: default
vignette: >
  %\VignetteIndexEntry{stats2data_application}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 6,
  fig.height = 4,
  fig.align  = "left",
  dpi        = 300
)
```

# Overview

This vignette walks through two empirical examples based on published studies whose raw data are not publicly available:

* an **ANOVA design** with reported cell means and F-statistics (Reynolds & Besner, 2008);
* a **multiple linear regression** with reported descriptives, correlations, and regression coefficients (Bardwell et al., 2007).

Across the two examples, every exported function in `stats2data` is demonstrated:

* the optimisation functions `optim_aov()`, `optim_vec()`, `optim_mlr()`, and `parallel_optim()`;
* the S3 inspection methods `print()`, `summary()`, `get_stats()`, `get_rmse()`, and `coef()`;
* the diagnostic and comparison plots `plot_error()`, `plot_cooling()`, `plot_summary()`, `plot()` (for parallel results), `plot_histogram()`, and `plot_partial_regression()`;
* the GRIM helper `check_grim()`;
* the Shiny entry point `run_app()`.

Tuning hyperparameters are left at their package defaults wherever feasible.

## Installation

```{r install, echo = TRUE, eval = FALSE}
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
devtools::install_github("sebastian-lortz/stats2data")
```

## Setup

```{r load}
library(stats2data)
library(gridExtra)
set.seed(2026)
```

# 1. ANOVA module --- Reynolds & Besner (2008)

In the context of the Open Science Collaboration's replication effort (Open Science Collaboration, 2015), the original data of Reynolds and Besner (2008, [https://osf.io/hasfu/](https://osf.io/hasfu/)) are not available. The authors reported a 2 × 2 within-subjects design measuring response times to exception words and nonwords under predictable switch and stay sequences. The reported summary statistics are encoded below as targets for `optim_aov()`.

## 1.1 Targets

```{r aov-targets}
S                  <- 16
levels             <- c(2, 2)
target_group_means <- c(543, 536, 614, 618)
factor_type        <- c("within", "within")
formula            <- "outcome ~ Factor1 * Factor2 + Error(ID / (Factor1 * Factor2))"
integer            <- FALSE
```

`Factor2` and the `Factor1:Factor2` interaction are reported as `F < 1`; placeholder targets are supplied for both:

```{r aov-fvalues}
target_f_list <- list(
  effect = c("Factor1", "Factor2", "Factor1:Factor2"),
  F_value      = c(30.5, .5, .5)
)
```

A plausible response-time range is derived from the reported pooled `MSE = 3070`:

```{r aov-range}
L <- min(target_group_means) - floor(4 * sqrt(3070))
U <- max(target_group_means) + ceiling(5 * sqrt(3070))
c(L = L, U = U)
range <- c(315, 896)
```

## 1.2 Run the optimiser

`optim_aov()` is called with default arguments (`max_iter = 1e4`, `thresh = 5e-3`, `init_temp = 1e-3`, `cooling_rate = NULL` (set automatically), `max_starts = 3`).

```{r aov-run}
result.aov <- optim_aov(
  S                  = S,
  levels             = levels,
  target_group_means = target_group_means,
  target_f_list      = target_f_list,
  factor_type        = factor_type,
  range              = range,
  formula            = formula,
  integer            = integer
  )
```

## 1.3 Inspect the result

The default `print` method shows the design, sample size, target effects, and best objective value:

```{r aov-print}
result.aov
```

`summary()` prints target-vs-simulated F-values, target-vs-simulated cell means, and per-block RMSE:

```{r aov-summary}
summary(result.aov)
```

`get_stats()` returns the underlying numeric components: the fitted `afex` ANOVA table, the F-vector, and the cell means computed from the simulated data.

```{r aov-stats}
stats.aov <- get_stats(result.aov)
stats.aov
```

`get_rmse()` returns the F-statistic and group-mean RMSE separately:

```{r aov-rmse}
get_rmse(result.aov)
```

## 1.4 Diagnostics

`plot_error()` shows the simulated-annealing best-error trajectory:

```{r aov-error, fig.alt = "Error trajectory for the ANOVA optimisation."}
plot_error(result.aov)
```

`plot_cooling()` reconstructs the temperature schedule from the stored `inputs`:

```{r aov-cooling, fig.alt = "Cooling schedule for the ANOVA optimisation."}
plot_cooling(result.aov)
```

## 1.5 Target vs. simulated comparison

`plot_summary()` builds a lollipop chart of simulated minus target values. With `standardised = FALSE` (used here because one target F is `0`), the y-axis is on the original F-scale; with `standardised = TRUE` (the package default) the discrepancy is divided by the target value, with a fallback to the unstandardised difference for near-zero targets.

```{r aov-summary-plot, fig.alt = "ANOVA target-vs-simulated lollipop chart, unstandardised."}
plot_summary(result.aov, standardised = FALSE)
```

## 1.6 Inspect the simulated dataset

The simulated dataset is stored at `result.aov$data` with columns `ID`, `Factor1`, `Factor2`, `outcome`:

```{r aov-data}
data.aov <- result.aov$data
head(data.aov)
```

We can vizualise the outcomes within groups with a density plot:

```{r aov-hist, fig.alt = "Distribution of the simulated outcome variable."}
library(ggplot2)

ggplot(data.aov,
       aes(x = outcome,
           fill = interaction(Factor1, Factor2, sep = " \u00d7 "))) +
  geom_density(alpha = 0.3, color = "gray20") +
  scale_fill_grey(start = 0.85, end = 0.25,
                  name = "Factor1 \u00d7 Factor2") +
  labs(x = "Outcome", y = "Density") +
  xlim(200,1000) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

# 2. Descriptives + MLR module --- Bardwell et al. (2007)

Bardwell et al. (2007, [https://doi.org/10.1016/j.jad.2006.06.013](https://doi.org/10.1016/j.jad.2006.06.013)) examined whether mood symptoms or apnea severity better predict daytime fatigue in patients with obstructive sleep apnea. The original raw data of Bardwell et al. (2003) are not publicly available. The reconstruction proceeds in three steps: marginal distributions via `optim_vec()`, joint regression structure via `optim_mlr()`, and a replication assessment via `parallel_optim()`.

## 2.1 Step 1 --- marginals via `optim_vec()`

### Targets

```{r vec-targets}
N           <- 60
target_mean <- c(48.8, 17.3, 12.6, 10.8)
names(target_mean) <- c("Apnea.1", "Apnea.2", "Depression", "Fatigue")
target_sd   <- c(27.1, 20.1, 11.3, 7.3)
integer     <- c(FALSE, FALSE, TRUE, TRUE)
range_matrix <- matrix(
  c( 15,   0,  0,  0,
    111, 80.9, 49, 28),
  nrow = 2, byrow = TRUE
)
```

### Sanity check with `check_grim()`

The GRIM test (Brown & Heathers, 2017) verifies whether a reported mean is achievable for the stated sample size and decimal precision. Each integer-valued mean is checked at one decimal place:

```{r vec-grim}
check_grim(n = N, target_mean = target_mean["Depression"], decimals = 1)
check_grim(n = N, target_mean = target_mean["Fatigue"],    decimals = 1)
```

### Run the optimiser

We set `sprite_prec = c(1, 1)`,`thresh = .005`, to set the reported rounding precision. Again, we run the optimization with default arguments use the defaults (`thresh = .005`, `max_iter = 5e5`, `init_temp = cooling_rate = NULL` (set automatically), `max_starts = 3`).

```{r vec-run}
result.vec <- optim_vec(
  N           = N,
  target_mean = target_mean,
  target_sd   = target_sd,
  range       = range_matrix,
  integer     = integer,
  sprite_prec = c(1, 1),
  thresh = .05
)
```

### Inspect

```{r vec-print}
result.vec
```

```{r vec-summary}
summary(result.vec)
```

```{r vec-stats}
get_stats(result.vec)
get_rmse(result.vec)
```

### Diagnostics

`plot_error()` for a `stats2data_vec` object accepts `run` to select a per-variable trajectory; here the trajectory for `Fatigue` (the fourth variable) is shown:

```{r vec-error, fig.alt = "Error trajectory for the Fatigue marginal."}
plot_error(result.vec, run = 2)
```

```{r vec-cooling, fig.alt = "Cooling schedule for the descriptives optimisation."}
plot_cooling(result.vec)
```

### Target vs. simulated comparison

```{r vec-summary-plot, fig.alt = "Descriptives target-vs-simulated lollipop chart."}
plot_summary(result.vec, standardised = FALSE)
```

### Inspect the simulated dataset

```{r vec-data, fig.width = 8, fig.height = 6, fig.alt = "Histograms of the four simulated marginals."}
data.vec <- result.vec$data
head(data.vec)

library(ggplot2)

cont_vars <- c("Apnea.1", "Apnea.2")
int_vars  <- c("Depression", "Fatigue")

ggdata.vec <- tidyr::pivot_longer(
  data.vec,
  cols      = c(Apnea.1, Apnea.2, Depression, Fatigue),
  names_to  = "variable",
  values_to = "value"
)
ggdata.vec$variable <- factor(
  ggdata.vec$variable,
  levels = c(cont_vars, int_vars)
)
ggdata.vec$kind <- ifelse(ggdata.vec$variable %in% int_vars,
                          "integer", "continuous")

ggplot(ggdata.vec, aes(x = value)) +
  geom_density(
    data  = subset(ggdata.vec, kind == "continuous"),
    fill  = "gray60", color = "gray20", alpha = 0.5
  ) +
  geom_histogram(
    data     = subset(ggdata.vec, kind == "integer"),
    aes(y    = after_stat(density)),
    binwidth = 1,
    fill     = "gray60", color = "gray20"
  ) +
  facet_wrap(~ variable, scales = "free", nrow = 2) +
  labs(x = NULL, y = "Density") +
  theme_minimal(base_size = 13)
```

## 2.2 Step 2 --- regression structure via `optim_mlr()`

### Targets

The reported regression coefficients (intercept first), their standard errors, and the upper-triangular correlation matrix (column-wise, excluding the diagonal) are encoded as numeric vectors. `NA` entries flag unreported values that are excluded from the objective.

```{r lm-targets}
target_reg <- c(4.020, 0.023, 0.008, 0.438)
names(target_reg) <- c("(Intercept)", "Apnea.1", "Apnea.2", "Depression")
target_se   <- c(NA, 0.034, 0.048, 0.066)
target_cor  <- c(NA, NA, NA, 0.11, 0.20, 0.68)
reg_equation <- "Fatigue ~ Apnea.1 + Apnea.2 + Depression"
```

### Run the optimiser

For compile-time reasons the local-refinement step is disabled (`hill_climbs = NULL`; default `1e4`), `max_starts` is reduced to `1` (default `3`), and `max_iter` is reduced to `1e4` (default `1e5`). All other arguments use the defaults (`weight = 0.5`, `thresh = 5e-3`, `init_temp = NULL` for adaptive initialisation, `cooling_rate = NULL`).

```{r lm-run}
result.mlr <- optim_mlr(
  N            = N,
  target_mean  = target_mean,
  target_sd    = target_sd,
  range        = range_matrix,
  integer      = integer,
  reg_equation = reg_equation,
  target_cor   = target_cor,
  target_reg   = target_reg,
  target_se    = target_se,
  thresh       = .0005,
  weight = .05,
)
```

### Inspect

```{r lm-print}
result.mlr
```

```{r lm-summary}
summary(result.mlr)
```

`get_stats()` returns the fitted `lm`, its coefficients and standard errors, the upper-triangular correlation vector, and the column-wise means and SDs:

```{r lm-stats}
stats.mlr <- get_stats(result.mlr)
stats.mlr$reg
stats.mlr$se
stats.mlr$cor
```

`coef()` is a convenience wrapper that re-fits the regression and returns the coefficient vector:

```{r lm-coef}
coef(result.mlr)
```

```{r lm-rmse}
get_rmse(result.mlr)
```

### Diagnostics

`plot_error()` shows the total-objective trajectory:

```{r lm-error, fig.alt = "Error trajectory for the MLR optimisation."}
plot_error(result.mlr, first_iter = 1000)
```

For `stats2data_mlr` objects, `plot_error()` exposes a `ratio = TRUE` switch that plots the per-iteration correlation-to-regression error ratio together with mean, median, and final reference lines:

```{r lm-error-ratio, fig.alt = "Cor/Reg error ratio trajectory for MLR."}
plot_error(result.mlr, ratio = TRUE)
```

We observe that the regression errors dominate the objective function, thus, we assign a priori more weight to the regression component compared to the correlation component by `weight = .05`. We run the optimization again, and plot the errror ratio:
```{r lm-cooling, fig.alt = "Weight tuned MLR optimisation."}
result.mlr <- optim_mlr(
  N            = N,
  target_mean  = target_mean,
  target_sd    = target_sd,
  range        = range_matrix,
  integer      = integer,
  reg_equation = reg_equation,
  target_cor   = target_cor,
  target_reg   = target_reg,
  target_se    = target_se,
  thresh       = .0005,
  weight = .05,
)
plot_error(result.mlr, ratio = TRUE)
```

By doing that we see a more balanced contribution of both components, that further lowered the objective value, increasing the match with the target summary statistics.

### Target vs. simulated comparison

```{r lm-summary-plot, fig.alt = "MLR target-vs-simulated lollipop chart."}
plot_summary(result.mlr, standardised = FALSE)
```

### Inspect the simulated dataset

```{r lm-data}
data.mlr <- result.mlr$data
head(data.mlr)
```

`plot_partial_regression()` returns one partial-regression ggplot per predictor, each annotated with the partial slope and the residual SD:

```{r lm-partial, fig.width = 11, fig.height = 8, fig.alt = "Partial-regression plots for the MLR fit."}
partial.plots <- plot_partial_regression(lm(reg_equation, data = data.mlr))
gridExtra::grid.arrange(grobs = partial.plots, ncol = 2)
```

## 2.3 Step 3 --- repeated runs via `parallel_optim()`

`parallel_optim()` repeatedly calls any `optim_*` function with the same arguments, returns the list of results, and identifies the single best run by minimum objective error. `runs = 18` parallel replications are used here for compile-time reasons; setting `cores = 6` dispatches the runs to multiple R sessions via `future::multisession`.

```{r lm-parallel, results = "hide", message = FALSE}
result.parallel <- parallel_optim(
  FUN = optim_mlr,
  args = list(
    N            = N,
    target_mean  = target_mean,
    target_sd    = target_sd,
    range        = range_matrix,
    integer      = integer,
    reg_equation = reg_equation,
    target_cor   = target_cor,
    target_reg   = target_reg,
    target_se    = target_se,
    thresh       = .0005,
    weight       = .05
  ),
  runs  = 18,
  cores = 6,
  seed  = 2026
)
```

### Inspect

```{r par-print}
result.parallel
```

```{r par-summary}
summary(result.parallel)
```

`get_stats()` for a `stats2data_parallel` object aggregates each numeric component returned by the per-run `get_stats()` method into a data frame with columns `mean`, `median`, `sd`, `min`, `max`:

```{r par-stats}
get_stats(result.parallel)
```

`get_rmse()` returns three components: `between_rmse` (per-metric run-to-run RMSE against the across-run grand mean), `target_rmse` (per-metric RMSE against the original targets), and the raw per-run vectors:

```{r par-rmse}
get_rmse(result.parallel)
```

### RMSE distribution

The `plot()` method for `stats2data_parallel` displays the between-run RMSE distribution alongside the deviation-from-target RMSE distribution, faceted by metric:

```{r par-plot, fig.width = 10, fig.height = 5, fig.alt = "Between-run vs. target RMSE distributions across parallel runs."}
plot_summary(result.parallel)
```

The single best run is accessible via `$best` and supports the same S3 methods as a stand-alone result:

```{r par-best}
result.parallel$best
get_rmse(result.parallel$best)
```

# 3. Shiny interface

A point-and-click interface to all three modules is launched via:

```{r app, eval = FALSE}
run_app()
```

# References
- Bardwell, W. A., Moore, P., Ancoli-Israel, S., & Dimsdale, J. E. (2003). Fatigue in obstructive sleep apnea: driven by depressive symptoms instead of apnea severity?. The American journal of psychiatry, 160(2), 350–355. https://doi.org/10.1176/appi.ajp.160.2.350
- Bardwell, W. A., Ancoli-Israel, S., & Dimsdale, J. E. (2007). Comparison of the effects of depressive symptoms and apnea severity on fatigue in patients with obstructive sleep apnea: a replication study. Journal of affective disorders, 97(1-3), 181–186. https://doi.org/10.1016/j.jad.2006.06.013
- Brown, N. J. L., & Heathers, J. A. J. (2017). The GRIM Test: A Simple Technique Detects
Numerous Anomalies in the Reporting of Results in Psychology. Social Psychological
and Personality Science, 8 (4), 363–369. https://doi.org/10.1177/1948550616673876
- Open Science Collaboration. (2015). Estimating the reproducibility of psychological science.
Science, 349 (6251), aac4716. https://doi.org/10.1126/science.aac4716
- Reynolds, M., & Besner, D. (2008). Contextual effects on reading aloud: Evidence for pathway control. Journal of Experimental Psychology: Learning, Memory, and Cognition, 34(1), 50–64. https://doi.org/10.1037/0278-7393.34.1.50
